Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-08-02 paper/paper.tex
p and q in vector format

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel distortion-corrected kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
\end{keywords}

@@

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this problem and measure the distortion incurred in manifold learning,
\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.

By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding. Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point. Then the DC\_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information. Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold. These density estimates are useful in many other areas, including classification, clustering and anomaly detection. Similar to \textcite{Cheng2021-ex}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection. \autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.

@@

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density esitmator to Riemnnian manifolds. However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional manifold learning embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by volume density function at each point. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds. Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}. Finally, we show how to use these densities to carrt out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}. Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathbb{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). To estimate the density of the manifold, \textcite{Pelletier2005-vu} propose a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation}
where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathbb{R}_{+} \rightarrow \mathbb{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\). The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.

Denote \(\lambda\) as the Lebesque measure of \(\mathbb{R}^d\). The kernels \(K\) in \eqref{eq:denriem} are functions of the geodesic distance of the manifold \(d_g\) and they satisfy the following conditions \autocite{Pelletier2005-vu}:
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(x)=K(0).
\end{align}
Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\). Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_pM\) centered at \(\pmb{p}\) with radius \(r\).

Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\). Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
\begin{equation}
\label{eq:volden}
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\end{equation}

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.

\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.

@@

By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates. When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates. In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_p(q)) |\).
In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding. Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid. Instead, we need a volume density function independent of the chart.

For \(p \in M\), denote an open ball centered at \(p\) with radius \(r\) as \(B(p,r)\) and the open neighborhood of \(0_p \in T_pM\) can be denoted as \(B(0_p,r)=\exp_p^{-1}(B(p,r))\). Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_p,r)\), with \(g^{\prime} = \exp^{\ast}_p(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_p\) in \(B(0_p,r)\).
Let \(u \in B(0_p,r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\), we have \(u \in \bar{U}\).
The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\). The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
Therefore, given a point \(q \in B(p,r)\) and a tangent vector \(u = \exp^{-1}_p(q) \in B(0_p,r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_p,r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
\begin{equation}
\label{eq:volumedensityratio}
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
\end{equation}
The volume density function is only defined for points satisfying \(d_g(p,q) < inj_gM\) so that there exists an open ball \(B(p,r) \in T_pM\). And \(\theta_p(q)=1\) for all \(p,q \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.\\
Then the volume density function is given by
\begin{equation}
\label{eq:volumedensityest}
\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
\end{equation}
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in Condition (v) of \eqref{eq:kernelcondition}.

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}. The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathcal{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathcal{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation. Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}. The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f(\pmb{y}_i)\).

>>>>

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
\end{keywords}

@@

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this problem and measure the distortion incurred in manifold learning,
\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.

By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding. Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point. Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information. Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold. These density estimates are useful in many other areas, including classification, clustering and anomaly detection. Similar to \textcite{Cheng2021-ex}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection. \autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.

@@

For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. \textcite{Pelletier2005-vu} generalize the multivariate kernel density estimator to Riemannian manifolds. However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \eqref{eq:denestimator0} based on the low-dimensional manifold learning embedding, \(\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), of the high-dimensional data,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\) is the pointwise adaptive Riemannian matrix of dimension \(d \times d\).
The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

To start with, we introduce the multivariate kernel density estimation method with variable bandwidth matrix and Pelletier's kernel density estimator for Riemannian manifolds. Then we propose our distortion-corrected estimator, followed by details about how to estimate the Riemannian metric using the Learn Metric algorithm from \textcite{Perrault-Joncas2013-pq}. Finally, we show how to use these densities to carry out anomaly detection tasks for high dimensional data that lies on Riemannian manifolds.

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) as a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}. Denote \(\pmb{X}_i \in \mathbb{R}^d, i = 1, 2, \dots, N\), as i.i.d. random objects on \(M\) with density \(f\). To estimate the density of the manifold, \textcite{Pelletier2005-vu} propose a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation}
where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathbb{R}_{+} \rightarrow \mathbb{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\). The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.

Denote \(\lambda\) as the Lebesque measure of \(\mathbb{R}^d\). The kernels \(K\) in \eqref{eq:denriem} are functions of the geodesic distance of the manifold \(d_g\) and they satisfy the following conditions \autocite{Pelletier2005-vu}:
\begin{align}
\label{eq:kernelcondition}
& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(x)=K(0).
\end{align}
Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\). Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_{\pmb{p}}M\) centered at \(\pmb{p}\) with radius \(r\).

Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
For any tangent vector \(u \in T_{\pmb{p}}M\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denote a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\). Then \(\frac{Dw_i}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_{\pmb{p}}M\).
The volume density function on \(M\), \(\theta_{\pmb{p}}(u): T_{\pmb{p}}M \rightarrow \mathbb{R}\) is defined in \textcite{Le_Brigant2019-lj} as
\begin{equation}
\label{eq:volden}
\theta_{\pmb{p}}(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\end{equation}

For the exponential map \(\exp_{\pmb{p}}: T_{\pmb{p}}M \rightarrow M\) induced by pullback of a volume form \(\exp_{\pmb{p}}^*\textit{Vol}\) on \(T_{\pmb{p}}M\), \(\theta_{\pmb{p}}\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_{\pmb{p}}M\), which gives
\[
d\exp_{\pmb{p}}^*\textit{Vo}l(u) = \theta_{\pmb{p}}(u)du,
\]
in normal coordinates \autocite{Le_Brigant2019-lj}. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. The volume form is variant to the exponential map except that the manifold is flat. Therefore, a multiplicative corrected term called the volume density function has to be added in the estimator so that it will integral to 1. It is closely related to the integration on manifolds and plays an important part in studying the density of manifolds.

\textcite{Pelletier2005-vu} mention that the volume density function \(\theta_{\pmb{p}}(\pmb{q})\) defined for \(\pmb{q}\) in a neighborhood of \(\pmb{p}\) on the manifold, with the geodesic normal coordinates at \(\pmb{p}\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of \(\pmb{q}\), \(\exp_{\pmb{p}}^{-1}(\pmb{q})\).
Note that this is true only when using the geodesic normal coordinates. If the mapping function is not exponential/logarithm map, the volume density function is calculated otherwise. Therefore, the Riemannian manifold density estimator in \textcite{Pelletier2005-vu} only works when the true manifold \(M\) is known.
Further, the fixed bandwidth assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point, which needs to be corrected using an adaptive bandwidth matrix when distortion exists.

@@

By the definition of the volume density function in Equation \eqref{eq:volden}, we could calculate how the volume change under the change of coordinates. When using the logarithm map, the coordinates of points on the manifold are transformed into their coordinates in the tangent space, i.e.~geodesic normal coordinates. In this case, \textcite{Pelletier2005-vu} gives the volume density function from the manifold to the tangent space, which is equal to the determinant of the Riemannian metric expressed in the geodesic normal coordinates at the logarithm map, \(|\det g(\exp^{-1}_{\pmb{p}}(\pmb{q})) |\).
In order to estimate the density of the true manifold, we first map the high-dimensional data to a low-dimensional embedding, then we apply a kernel density estimator to the embedding. Since the logarithm map of the manifold is also unknown, the volume density function in \textcite{Pelletier2005-vu} is no longer valid. Instead, we need a volume density function independent of the chart.

For \(\pmb{p} \in M\), denote an open ball centered at \(\pmb{p}\) with radius \(r\) as \(B(\pmb{p},r)\) and the open neighborhood of \(0_{\pmb{p}} \in T_{\pmb{p}}M\) can be denoted as \(B(0_{\pmb{p}},r)=\exp_{\pmb{p}}^{-1}(B(\pmb{p},r))\). Consider the Riemannian metrics \(g^{\prime}\) and \(g^{\prime \prime}\) in \(B(0_{\pmb{p}},r)\), with \(g^{\prime} = \exp^{\ast}_{\pmb{p}}(g)\) being the pullback of \(g\) induced by the exponential map and \(g^{\prime \prime}\) being the canonical metric induced by \(g_{\pmb{p}}\) in \(B(0_{\pmb{p}},r)\).
Let \(u \in B(0_{\pmb{p}},r)\) be a tangent vector and for any given chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\), we have \(u \in \bar{U}\).
The volumes of the parallelepiped spanned by the basis \(\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}\) are given by \(| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime}\), and \(| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}\) with respect to \(g^{\prime \prime}\). The quotient of these two volumes is then independent of the chart \((\bar{U}, \bar{\psi})\) \autocite{Henry2009-om}.
Therefore, given a point \(\pmb{q} \in B(\pmb{p},r)\) and a tangent vector \(u = \exp^{-1}_{\pmb{p}}(\pmb{q}) \in B(0_{\pmb{p}},r)\), for any chart \((\bar{U}, \bar{\psi})\) of \(B(0_{\pmb{p}},r)\) that contains \(u\), \textcite{Henry2009-om} defines the volume density function on the Riemannian manifold \((M,g)\) as
\begin{equation}
\label{eq:volumedensityratio}
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
\end{equation}
The volume density function is only defined for points satisfying \(d_g(\pmb{p},\pmb{q}) < inj_gM\) so that there exists an open ball \(B(\pmb{p},r) \in T_{\pmb{p}}M\). And \(\theta_{\pmb{p}}(\pmb{q})=1\) for all \(\pmb{p},\pmb{q} \in \mathbb{R}^d\) when the manifold \(M\) is \(\mathbb{R}^d\) with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart \((U, \psi)\) induced by an orthonormal basis of the tangent space of the embedding \(T_{\pmb{y}_i}E\) and a normal neighborhood of \(\pmb{y}_i\). Given a query point \(\pmb{p} \in B(\pmb{y}_i,r)\), the volume of the parallelepiped is spanned by the same basis.
Then the volume density function is given by
\begin{equation}
\label{eq:volumedensityest}
\theta_{\pmb{y}_i}(\pmb{p}) = \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2},
\end{equation}
where \(\pmb{H}\) is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all \(\pmb{p}\) but for points within the geodesic ball satisfying \(\pmb{p} \in B(\pmb{y}_i,r)\), which is related to the support of \(K\) being \([0,1]\) in Condition (v) of \eqref{eq:kernelcondition}.

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \eqref{eq:denestimator}. The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \ref{alg:kdeanomaly}.
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \ref{alg:learnmetric} to the input high-dimensional data \(x_i,i=1,\ldots,N\) where \(x_i\in \mathbb{R^s}\) to get the low-dimensional embedding \(y_i\), where \(y_i\in \mathbb{R^d}\) and the Riemannian metric \(\pmb{H}_i\) with dimension \(d\times d\) for each observation. Two parameters \(\sqrt{\varepsilon} = 0.4\) and \(c=0.25\) are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \eqref{eq:denestimator}. The top outliers of size \(n\_outliers\) are obtained by ordering the embedding points \(y_i\) according to their density estimates \(f(\pmb{y}_i)\).
